Investigating the impact of interventionist causal approach on the study of verbal aggression online

### 2021-09-09 paperDraft/paperDraft.tex
changes made to philosophical grounding of intevrentions

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating the impact of interventionist causal approach on the study of verbal aggression and discrimination online},
  pdfauthor={Patrycja Tempska and Rafal Urbaniak},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\documentclass{article}

@@

\title{Investigating the impact of interventionist causal approach on
the study of verbal aggression and discrimination online}
\author{Patrycja Tempska and Rafal Urbaniak}
\date{}

@@

\hypertarget{section-abstract}{%
\section{Section Abstract}\label{section-abstract}}

This article describes an experimental intervention study conducted in a
naturalistic, digital setting (Q\&A forum on Reddit), utilizing a
collective intelligence approach to content moderation and reduction of
the level of verbal aggression among a selected group of Reddit users
who regularly attack other community members. Collective Intelligence in
this sense means exploring the collaboration between human and machine
intelligence to develop solutions to social challenges. Artificial
Intelligence was used to detect verbal aggression (personal attacks) and
notify human volunteers about attacks. Volunteers after receiving
notifications employed interventions based on norm or empathy promotion.
We find that only those who were sanctioned with norms-inducing
interventions had their personal attacks' level significantly decreased.

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Although much effort has been made in order to tackle the problem of
verbal aggression and harassment online, looking at various reports and
surveys (Laub, 2019; Sorrentino, Baldry, Farrington, \& Blaya, 2019;
Vogels, 2021), it remains a common hindrance for people engaging with
social media in their everyday lives. The problem got exacerbated in the
midst of the COVID19 pandemic, during which the majority of our social
life moved to cyberspace. During this shift, there was an increase in
cyberbullying attitudes and perpetration (Barlett, Simmers, Roth, \&
Gentile, 2021), 90\% increase in public reports of illegal online
content (Grant, 2021), including 114\% increase in non-consensual
sharing of intimate images, 30\% increase in cyberbullying, as well as
40\% of increase in adults reporting online harassment. According to a
report conducted by company L1ght (L1ght, 2020), hate speech directed
towards China and the Chinese went up by 900\% on Twitter. Gaming
platforms were in the spotlight as well, with a 40\% increase in
toxicity on Discord
\footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

But alongside the growing need for even more efficient and proactive
moderation, the capacity to execute it did not go hand in hand, forcing
companies and policymakers to rethink the current model of moderation
processes and workforce. Due to the COVID19 restrictions including
social distancing, a lot of those serving the role of moderators had to
be sent home (Bhattacharya, 2021) without the ability to work remotely
because of the constraints affiliated with restrictive non-disclosure
agreements (NDA) among others. Curtailing the moderators' workforce was
accompanied by more agency given to algorithms and AI-based moderation.
Those changes, as argued by Gerrard (2020), can be seen as a serious red
flag in terms of safety for all users on online platforms.

@@

Traditionally collective intelligence has been defined as ``a group or a
team's combined capacity and capability to perform a wide variety of
tasks and solve diverse problems'' ({``Collective {Intelligence},''}
n.d.). In our paper and in the theoretical underpinnings of the
experiment itself, we will be relying on a collective intelligence scope
proposed by Nesta, an innovation foundation
(\url{https://www.nesta.org.uk}), which focuses on a collaboration
between human and machine intelligence to develop innovative solutions
to social challenges. (Gdzieś stopka(?): Samurai Labs has been one of
the 15 recipients of the second round of collective intelligence grants
awarded by Nesta:
\url{https://www.nesta.org.uk/project-updates/second-round-collective-intelligence-grants/})

@@

\textbf{Empathetic interventions} In the first treatment condition,
volunteers were encouraged to send empathy-inducing messages focusing
either on the target of verbal aggression (e.g.~„Hey such words might
hurt the other person''), stressing the common humanity aspect that we
call share („We are all humans of flesh and blood''), or even infusing
the intervention with the emphatic response to the attacker (Hey I
understand your strong emotions\ldots''). At the core of the empathetic
interventions, we put forth the notion that goes back to David Hume and
Adam Smith. The first one conceived empathy (at the time referred to as
sympathy) as mirroring the emotional state of another person. In the
academic psychological literature, similar phenomenon was distinguished
and coined in the term emotional contagion
(\url{https://www.sciencedirect.com/topics/psychology/emotional-contagion}):
``the process in which an observed behavioral change in one individual
leads to the reflexive production of the same behavior by other
individuals in close proximity, with the likely outcome of converging
emotionally.'' For Adam Smith, sympathy consisted of visualizing how the
sympathetic person would feel in the particular circumstances of the
other --- thus here the process was based not so much on mirroring, but
rather projecting my imagination of what it is like to be that person in
a certain moment). Without further delving into the differentiation
between those two, sympathy in both accounts is crucial in the
constitution of human beings as social and moral creatures
(\url{https://plato.stanford.edu/entries/empathy/}). It enables the
emotional connection to others and concern for their well-being. In the
psychological literature, various kinds of empathetic responses were
distinguished - the aforementioned emotional contagion, affective/proper
empathy, sympathy, personal distress or cognitive empathy
((\url{https://plato.stanford.edu/entries/empathy/}). During the
experiment, we couldn't observe or measure whether the empathetic
response indeed was evoked. Also, interventions that stated that ``such
words might hurt the other person'' have the underlying assumption that
the receiver of the attack might be hurt, but in reality that might not
even be the case. Intuitively, and following the argumentation of Thomas
Nagel in ``What is it like to be a bat?'' one can only imagine what it
is like - but for me - to be a bat or to be a receiver of the attack. To
each observed or imagined experience there is an array of subjective
quality to actually experiencing it and in this way receiving a
particular message might be met with a unique reception by each
conscious being. Also, full epistemic access to the mind and body of
another is impossible. But even though such access is impossible and
putting aside the broad spectrum of phenomena related to empathetic
response, our goal was by utilizing interventions referring to empathy
in various forms, changing the behavior of the attacker - convincing him
to refrain from using verbal aggression towards others. Whether the
message indeed gave rise to empathy was not of importance - one can
imagine a hypothetical scenario of a successful intervention in which
the attacker changed his behavior long-term and stopped attacking
others, hopefully as a result of interventions, but the empathetic
response was not even evoked. The behavioral change could be induced by
other motivations - e.g.~unwillingness from the exclusion from the
community. Thus the goal is not to evoke empathy, but through empathetic
interventions - substantially limiting the use of verbal aggression.

\textbf{Normative interventions} Normative interventions were grounded
in broadly defined norms and normative theories expressed via various
means - either specific community guidelines imposed by Reddit or
through ethics - deontology, virtue ethics. Whatever the strategy was
used, it was supposed to express the unacceptability of verbal
aggression in a direct or non-direct way.

A more direct approach referred to deontology in which „actions are good
or bad according to a clear set of rules''
(\url{https://ethics.org.au/ethics-explainer-deontology/}). Such
interventions could state that „we have a duty to respect each other
during the discussions``, „we have a moral obligation to act in a civil
manner while participating in online communities,'' „we shouldn't use ad
personam in here''. The basic assumption behind those interventions is
that there is something we ought to do or are morally required to do -
and such actions are the right actions. There are also actions we ought
not to do - and such actions are morally wrongful. This line of thinking
can be traced back to Immanuel Kant who thought that all universal moral
obligations can stem from categorical imperative: ``act only in
accordance with that maxim through which you can at the same time will
that it become a universal law'' (Groundwork of the Metaphysic of
Morals).

Yet other interventions were encouraged to be expressed in the light of
virtue ethics and stressed the importance of adhering to or practicing
certain virtues. Such a message was assumed to motivate the moral agent
in the process of positive reinforcement and create a feeling of the
desirability of certain behaviors, e.g.~„capacity to be respectful in a
heated discussion is a virtue and requires hard work''. The core
assumption here is that the virtue itself is not genetically inherited
but is rather a potential or a disposition of a character that can be
practice and mastered like a practical skill (even though as mentioned
by Natasza Szutta in {[}cite her book about virtue ethics{]} virtues and
practical skills share some fundamental differences). NS distinguished
two types of virtues - affective and cognitive. Affective virtues
encompass the emotions and feelings that play an important and positive
role in morality and can act as a support in the course of becoming a
virtuous man. A cognitive virtue relates to the intellectual aspect in
which one knows how to act in certain situations and understands the
rules of morally rightful actions (e.g.~what is kindness and how a kind
man acts). Here, just as in the case of empathetic interventions, any
attempts to measure whether the target of the intervention in the case
of a positive outcome (behavioral change) indeed acted in a virtuous
way. A virtue of kindness may manifest itself in particular behaviors
but as such cannot be identified with the virtuous deed, as highlighted
by Natasza Szutta. Although again, the goal of the experiment itself was
to change the behavior of the attackers, and measuring whether
particular messages contributed to more flourishing individuals in terms
of virtue development and character creation lies beyond the scope of
this work. Interventions could affect people either way - stimulate
their moral reflections or limit their attacks due to consequentialist
way of thinking and fear of potential outcomes (e.g.~being excluded from
the community).

Additionally, Reddit created their social etiquette called
``Rediquette''
(\url{https://www.reddithelp.com/hc/en-us/articles/205926439}) which is
``an informal expression of the values of many redditors, as written by
redditors themselves.'' All users are encouraged to abide by it and
moderators of communities (called subreddits) existing as a part of the
platform have the authority to exclude its members based either on the
basis of breaking the rules of the reddiquette or any other local rules
imposed by specific subreddit. Volunteers were encouraged to refer to
those norms as well, citing specific points, for instance, ``Hey there,
have you read the reddiquette? It says remember the human.'' or ``Let's
recall the reddiquette and adhere to the same standards of behavior
online that you follow in real life.''

As we have seen, those two categories of interventions - normative and
empathetic - encompass a whole variety of categories within. One might
think that it would be useful to construct a more diverse set of
treatment conditions in which we test each of the ethics or empathy
differentiation separately - e.g.~one for virtue ethics, deontology,
perhaps utilitarian approach, empathy toward the receiver of the attack,
empathy towards the attacker, so on and so forth. Additionally,
different types of interventions could be tested and tailored depending
on the spectrum on which one can be found in the foundations of moral
reasoning, as proposed by Jonathan Haidt: Care/Harm, Fairness/Cheating,
Loyalty/Betrayal, Authority/Subversion, Sanctity/Degradation, and
Liberty/Oppression (J. Haidt, The Righteous Mind). Albeit for the sake
of the sanity of our volunteers, we decided to create only two but broad
categories and let them decide based on the context of the conversation,
how they want to respond to the attacker.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating the impact of interventionist causal approach on the study of verbal aggression online},
  pdfauthor={Patrycja Tempska and Rafal Urbaniak},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\documentclass{article}

@@

\title{Investigating the impact of interventionist causal approach on
the study of verbal aggression online}
\author{Patrycja Tempska and Rafal Urbaniak}
\date{}

@@

@@

\textbackslash section Introduction

We live in a world in which there is an abundance of data. Referring the
phrase coined by Clive Humby, ``data is the new oil'' --- data has
displaced oil and become the world's most valuable resource. Albeit
there are some fundamental differences between the two assets. Oil
belongs to a very limited and finite resource, while data is growing at
an exponential rate (cite:
\url{https://cloudtweaks.com/2015/03/how-much-data-is-produced-every-day/}).
The value of oil does not depend on its amount or the refinery processes
- it's rather constant at a particular dynamic of market's supply and
demand. Data and its value, on the contrary, is inherently tied to its
amount and the ``refinery process'' - various tools and techniques for
its discovery and analysis. In some cases, certain techniques can be
used only if one has a sufficient (meaning enormous) data at one's
disposal --- e.g.~statistical methods like machine learning and deep
learning \todo{@article{lecun2015deep,
  title={Deep learning},
  author={LeCun, Yann and Bengio, Yoshua and Hinton, Geoffrey},
  journal={nature},
  volume={521},
  number={7553},
  pages={436--444},
  year={2015},
  publisher={Nature Publishing Group}
}}. In the last couple of years, more and more industry and academia
representatives alike started to research and develop tools for causal
discovery and inference. Unfortunately, although thinking in causal
terms seems to be our inherent human capacity and one of our basic
cognitive abilities, that is not necessarily the case with the majority
of toolkits and statistical models developed to draw insight and
correlations from data. In classical statistics, which is stripped of
causality-related terms and reasoning, one cannot speak about causal
effects, unless one deals with a randomized controlled experiment. This
comprises one of the limits of the classical paradigm in statistics
(among others such as arbitrary thresholds of statistical significance,
p-hacking, disregard of prior information, and others which we will not
discuss here). But often scientists conduct the so-called
cross-sectional research, in which no variable is manipulated. In the
language and toolkit of classical statistics in such cases we cannot
conduct any causal inference. What we can measure using those methods is
correlation, and, as frequently repeated, ``correlation does not imply
causation.'' Correlation can be positive, negative, or there can be no
correlation at all. A positive one implies that the increase of the
value of one variable will influence the increase of another variables'
value. A negative correlation means that the change in values of two
variables will diverge into separate directions --- one will increase
and one will decrease alongside the first one. We can measure the
strength of the correlation using various mathematical methods, where 1
or -1 means we are dealing with perfect correlation --- the value of one
variable changes proportional to the value of another variable. But even
though causation is necessarily connected with correlation, as already
mentioned --- not every correlation implies causation. This is
especially problematic if we take into consideration the fact that
statisticians and people in general naturally think in causal terms and
spurious interpretations of studies occur frequently. On the other hand,
a conservative approach of many classical statisticians led to a
long-living debate on whether smoking causes cancer, lasting for over 20
years. One of the reasons for such a fierce debate was the lack of a
causal framework, which was still in its infancy at that time. Framework
of tools with which we could validate or invalidate assumptions about
the existence of the potential confounder --- a variable causing both
craving cigarettes and lung cancer (e.g.~some genetic factor), creating
a spurious correlation between cigarettes and lung cancer. Modeling
causality and moving beyond traditional statistics is a relatively young
endeavor. One of the pioneers of the work on causality is Judea Pearl.
As he describes, causality and traditional statistics are governed by
opposing ways of thinking. While statistics might be described as
``study of methods of the reduction of data,'' causality requires the
incorporation of an extra layer --- the understanding or prior knowledge
(assumptions) about the process that generate the data in the first
place. Classical statistics might be described here as a ``model-free
approach,'' just as Artificial Intelligence methods based on machine
learning/deep learning, in which no prior knowledge can be taken into
account.

In this paper, I will show how mathematical methods for process modeling
which includes assumptions about causal connections can be applied to
discover how different variables affect each other, moving beyond
classical statistics. Looking at the analysis from two perspectives ---
traditional (classical, frequentist statistics) and bayesian -- we will
see the advantage of the second paradigm in learning about the world
based on the data we have at our disposal. Data used for analysis
concerned in this paper comes from an experimental intervention study
conducted in a naturalistic, digital setting (Q\&A forum on Reddit),
utilizing a collective intelligence approach to content moderation and
reduction of the level of verbal aggression among a selected group of
Reddit users who regularly attack other community members. Collective
Intelligence in this sense means exploring the collaboration between
human and machine intelligence to develop solutions to social
challenges. Artificial Intelligence was used to detect verbal aggression
(personal attacks) and notify human volunteers about attacks. Volunteers
after receiving notifications employed interventions inspired in its
underlying principles in philosophy and psychology. There were two broad
categories of interventions - one based on norms (deontology, virtue
ethics among others) and the second based on empathy (inspired by
notions of David Hume and Adam Smith).

One might also wonder about the motivations for conducting such an
experiment. Much effort has been made in order to tackle the problem of
verbal aggression and harassment online, but looking at various reports
and surveys (Laub, 2019; Sorrentino, Baldry, Farrington, \& Blaya, 2019;
Vogels, 2021), it remains a common hindrance for people engaging with
social media in their everyday lives. The problem got exacerbated in the
midst of the COVID19 pandemic, during which the majority of our social
life moved to cyberspace. During this shift, there was an increase in
cyberbullying attitudes and perpetration (Barlett, Simmers, Roth, \&
Gentile, 2021), 90\% increase in public reports of illegal online
content (Grant, 2021), including 114\% increase in non-consensual
sharing of intimate images, 30\% increase in cyberbullying, as well as
40\% of increase in adults reporting online harassment. According to a
report conducted by company L1ght (L1ght, 2020), hate speech directed
towards China and the Chinese went up by 900\% on Twitter. Gaming
platforms were in the spotlight as well, with a 40\% increase in
toxicity on Discord
\footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

But alongside the growing need for even more efficient and proactive
moderation, the capacity to execute it did not go hand in hand, forcing
companies and policymakers to rethink the current model of moderation
processes and workforce. Due to the COVID19 restrictions including
social distancing, a lot of those serving the role of moderators had to
be sent home (Bhattacharya, 2021) without the ability to work remotely
because of the constraints affiliated with restrictive non-disclosure
agreements (NDA) among others. Curtailing the moderators' workforce was
accompanied by more agency given to algorithms and AI-based moderation.
Those changes, as argued by Gerrard (2020), can be seen as a serious red
flag in terms of safety for all users on online platforms. That is why
in the paper I will also try to sketch a landscape of current moderation
techniques through a critical lens.

@@

Traditionally collective intelligence has been defined as ``a group or a
team's combined capacity and capability to perform a wide variety of
tasks and solve diverse problems'' ({``Collective {Intelligence},''}
n.d.). In our paper and in the theoretical underpinnings of the
experiment itself, we will be relying on a collective intelligence scope
proposed by Nesta, an innovation foundation
(\url{https://www.nesta.org.uk}), which focuses on a collaboration
between human and machine intelligence to develop innovative solutions
to social challenges.

@@

\textbf{Empathetic interventions} In the first treatment condition,
volunteers were asked to send empathy-inducing messages focusing either
on the target of verbal aggression (e.g.~„Hey such words might hurt the
other person''), stressing the common humanity aspect that we call share
(„We are all humans of flesh and blood''), or even infusing the
intervention with the emphatic response to the attacker (Hey I
understand your strong emotions\ldots''). At the core of the empathetic
interventions, we put forth the notion that goes back to David Hume and
Adam Smith. The first one conceived empathy (at the time referred to as
sympathy) as mirroring the emotional state of another person. In the
academic psychological literature, similar phenomenon was distinguished
and coined in the term emotional contagion
(\url{https://www.sciencedirect.com/topics/psychology/emotional-contagion}):
``the process in which an observed behavioral change in one individual
leads to the reflexive production of the same behavior by other
individuals in close proximity, with the likely outcome of converging
emotionally.''

For Adam Smith, sympathy consisted of visualizing how the sympathetic
person would feel in the particular circumstances of the other --- thus
here the process was based not so much on mirroring, but rather
projecting my imagination of what it is like to be that person in a
certain moment). Without further delving into the differentiation
between those two, sympathy in both accounts is crucial in the
constitution of human beings as social and moral creatures
(\url{https://plato.stanford.edu/entries/empathy/}). It enables the
emotional connection to others and concern for their well-being. The
interventions were supposed to serve the role of empathetic response
enablers --- reminding the person in front of the screen that there is a
real human being on the other side.

In the psychological literature, various kinds of empathetic responses
were distinguished - the aforementioned emotional contagion,
affective/proper empathy, sympathy, personal distress or cognitive
empathy ((\url{https://plato.stanford.edu/entries/empathy/}). During the
experiment, we could not observe or measure whether the empathetic
response indeed was evoked. The study was conducted in a digital setting
--- interventions were most likely read in the private space, in front
of personal computers or smartphones. Even if we would change the
setting to a lab, differentiating and measuring empathy is not trivial
if not impossible (to check).

Also, interventions that stated that ``such words might hurt the other
person'' have the underlying assumption that the receiver of the attack
might be hurt, but in reality that might not even be the case.
Intuitively, and following the argumentation of Thomas Nagel in ``What
is it like to be a bat?'' one can only imagine what it is like --- but
for me --- to be a bat or to be a receiver of the attack. To each
observed or imagined experience there is an array of subjective quality
to actually experiencing it and in this way receiving a particular
message might be met with a unique reception by each conscious being.
Also, full epistemic access to the mind and body of another is
impossible.

But even though such access is impossible and putting aside the broad
spectrum of phenomena related to empathetic response, our goal was by
utilizing interventions referring to empathy in various forms, changing
the behavior of the attacker --- convincing him to refrain from using
verbal aggression towards others. Whether the message indeed gave rise
to empathy was not of importance --- one can imagine a hypothetical
scenario of a successful intervention in which the attacker changed his
behavior long-term and stopped attacking others, hopefully as a result
of interventions, but the empathetic response was not even evoked. The
behavioral change could be induced by other motivations ---
e.g.~unwillingness from the exclusion from the community. Thus the goal
is not to evoke empathy, but through empathetic interventions ---
substantially limit the use of verbal aggression among study group
members.

\textbf{Normative interventions} In the second treatment condition,
volunteers were asked to send norms-inducing messages written on the
grounds of deontology (e.g.~„we have a duty to respect each other during
the discussions``), virtue ethics (e.g.~„capacity to be respectful in a
heated discussion is a virtue and requires hard work'') or Reddit's
community guideline (e.g.~``Let's recall the reddiquette and adhere to
the same standards of behavior online that you follow in real life'').

Whatever the strategy was used, it was supposed to express the
unacceptability of verbal aggression in a direct or non-direct way. A
more direct approach referred to deontology in which „actions are good
or bad according to a clear set of rules''
(\url{https://ethics.org.au/ethics-explainer-deontology/}). The basic
assumption behind those interventions is that there is something we
ought to do or are morally required to do --- and such actions are the
right actions. There are also actions we ought not to do --- and such
actions are morally wrongful. This line of thinking can be traced back
to Immanuel Kant who thought that all universal moral obligations can
stem from categorical imperative: ``act only in accordance with that
maxim through which you can at the same time will that it become a
universal law'' (Groundwork of the Metaphysic of Morals).

In the psychological literature, Robert B. Cialdini distinguished three
types of norms that he found to be effective in guiding human action:
descriptive norms (focused on the perception of how the majority of
people would behave), injunctive (based on the perception of how the
majority of people would approve or disapprove certain conduct), or
personal norms (based on the perception of how a particular person would
approve or disapprove certain behavior) CITE:
\url{https://www.sciencedirect.com/science/article/abs/pii/S0065260108603305}.
Although interventions written on the grounds of duties and deontology
cannot be reduced to any of the above, since following Hume's guillotine
--- from the fact that the majority of people are respectful does not
follow the normative statement of civility --- I hoped that expressing a
universal normative statement will be even more powerful and potentially
spawn an interpretation of social acceptability or unacceptability of
certain behavior.

Yet other interventions in normative group were encouraged to be
expressed in the light of virtue ethics and stressed the importance of
adhering to or practicing certain virtues. Such a message was assumed to
motivate the moral agent in the process of positive reinforcement and
create a feeling of the desirability of certain behaviors, like kindness
and civility. The core assumption here is that the virtue itself is not
genetically inherited but is rather a potential or a disposition of a
character that can be practice and mastered like a practical skill (even
though as mentioned by Natasza Szutta in {[}cite her book about virtue
ethics{]} virtues and practical skills share some fundamental
differences). NS distinguished two types of virtues - affective and
cognitive. Affective virtues encompass the emotions and feelings that
play an important and positive role in morality and can act as a support
in the course of becoming a virtuous man. A cognitive virtue relates to
the intellectual aspect in which one knows how to act in certain
situations and understands the rules of morally rightful actions
(e.g.~what is kindness and how a kind man acts).

Here, just as in the case of empathetic interventions, any attempts to
measure whether the target of the intervention in the case of a positive
outcome (behavioral change) indeed acted in a virtuous way. A virtue of
kindness may manifest itself in particular behaviors but as such cannot
be identified with the virtuous deed, as highlighted by Natasza Szutta.
Although again, the goal of the experiment itself was to change the
behavior of the attackers, and measuring whether particular messages
contributed to more flourishing individuals in terms of virtue
development and character creation lies beyond the scope of this work.
Interventions could affect people either way - stimulate their moral
reflections or limit their attacks due to consequentialist way of
thinking and fear of potential outcomes (e.g.~being excluded from the
community).

Additionally, Reddit created their social etiquette called
``Rediquette''
(\url{https://www.reddithelp.com/hc/en-us/articles/205926439}) which is
``an informal expression of the values of many redditors, as written by
redditors themselves.'' All users are encouraged to abide by it and
moderators of communities (called subreddits) existing as a part of the
platform have the authority to exclude its members based either on the
basis of breaking the rules of the reddiquette or any other local rules
imposed by specific subreddit. Volunteers were encouraged to refer to
those norms as well, citing specific points, for instance, ``Hey there,
have you read the reddiquette? It says remember the human.''

As we have seen, those two categories of interventions - normative and
empathetic - encompass a whole variety of categories within. One might
think that it would be useful to construct a more diverse set of
treatment conditions in which we test each of the ethics or empathy
differentiation separately - e.g.~one for virtue ethics, deontology,
perhaps utilitarian approach, empathy toward the receiver of the attack,
empathy towards the attacker, so on and so forth. Additionally,
different types of interventions could be tested and tailored depending
on the spectrum on which one can be found in the foundations of moral
reasoning, as proposed by Jonathan Haidt: Care/Harm, Fairness/Cheating,
Loyalty/Betrayal, Authority/Subversion, Sanctity/Degradation, and
Liberty/Oppression (J. Haidt, The Righteous Mind). Albeit for the sake
of the sanity of our volunteers, we decided to create only two but broad
categories and let volunteers decide based on the context of the
conversation, how they want to respond to the attacker.
